---
title: Epidemiological analysis with R
author: Sergio Uribe
date: '2020-07-24'
slug: epidemiological-analysis-with-r
categories:
  - epidemiological research
tags:
  - EDA
  - epidemiology
subtitle: 'Riga Stradins University Workshop, July 2020'
---
+ Lecturer: Sergio Uribe, Assoc Prof, PhD
+ Riga Stradins University, July 2020


Required packages

```{r}
pacman::p_load(
  car,
  broom,
  tidyverse,
  ggfortify,
  mosaic,
  huxtable,
  jtools,
  latex2exp,
  pubh,
  sjlabelled,
  sjPlot,
  sjmisc
)

theme_set(sjPlot::theme_sjplot2(base_size = 10))
theme_update(legend.position = "top")
# options('huxtable.knit_print_df' = FALSE)
options('huxtable.knit_print_df_theme' = theme_article)
options('huxtable.autoformat_number_format' = list(numeric = "%5.2f"))
knitr::opts_chunk$set(collapse = TRUE, comment = NA)
```

# Epidemiological Descriptive Analysis 

Onchocerciasis in Sierra Leone.
```{r}
data(Oncho)
Oncho %>% head()
```
A two-by-two contingency table:
```{r}
Oncho %>%
  mutate(mf = relevel(mf, ref = "Infected")) %>%
  copy_labels(Oncho) %>%
  cross_tab(mf ~ area) 
```

Table with all descriptive statistics except id and mfload:

```{r}
Oncho %>%
  # dplyr:select(-c(id, mfload)) %>%
  dplyr::select(-id) %>%
  dplyr::select(-mfload) %>%  # just for example
  mutate(mf = relevel(mf, ref = "Infected")) %>%
  copy_labels(Oncho) %>%
  cross_tab(mf ~ area + .) 
```

Data set about blood counts of T cells from patients in remission from Hodgkin's disease or in remission from disseminated malignancies. We generate the new variable Ratio and add labels to the variables.
```{r}
data(Hodgkin)
Hodgkin <- Hodgkin %>%
  mutate(Ratio = CD4/CD8) %>%
  var_labels(
    Ratio = "CD4+ / CD8+ T-cells ratio"
    )

Hodgkin %>% head()
```

Descriptive statistics for CD4+ T cells:
```{r}
Hodgkin %>%
  estat(~ CD4)
```
For stratification, estat recognises the following two syntaxes:

 - outcome ~ exposure
 - ~ outcome | exposure
 
where, outcome is continuous and exposure is a categorical (factor) variable.

```{r}
Hodgkin %>%
  estat(~ Ratio|Group)
```

descriptive statistics for all variables in the data set:

```{r}
Hodgkin %>%
  mutate(Group = relevel(Group, ref = "Hodgkin")) %>%
  copy_labels(Hodgkin) %>%
  cross_tab(Group ~ CD4 + .)
  # theme_article() %>%
  # add_footnote("Values are medians with interquartile range.")
```
## Inferential statistics

```{r}
var.test(Ratio ~ Group, data = Hodgkin)
```

QQ-plots against the standard Normal distribution.
```{r}
Hodgkin %>%
  qq_plot(~ Ratio|Group) %>%
  axis_labs()
```
non-parametric test to compare the groups:
```{r}
wilcox.test(Ratio ~ Group, data = Hodgkin)
```

For relatively small samples (for example, less than 30 observations per group) is a standard practice to show the actual data in dot plots with error bars. The pubh package offers two options to show graphically differences in continuous outcomes among groups:

For small samples: strip_error
For medium to large samples: bar_error

```{r}
Hodgkin %>%
  strip_error(Ratio ~ Group) %>%
  axis_labs()
```

In the previous code, axis_labs applies labels from labelled data to the axis.

The error bars represent 95% confidence intervals around mean values.

Is relatively easy to add a line on top, to show that the two groups are significantly different. The function gf_star needs the reference point on how to draw an horizontal line to display statistical difference or to annotate a plot; in summary, gf_star:

```{r}
Hodgkin %>%
  strip_error(Ratio ~ Group) %>%
  axis_labs() %>%
  gf_star(x1 = 1, y1 = 4, x2 = 2, y2 = 4.05, y3 = 4.1, "**")
```

For larger samples we could use bar charts with error bars. For example:

```{r}
data(birthwt, package = "MASS")
birthwt <- birthwt %>%
  mutate(
    smoke = factor(smoke, labels = c("Non-smoker", "Smoker")),
    Race = factor(race > 1, labels = c("White", "Non-white")),
    race = factor(race, labels = c("White", "Afican American", "Other"))
    ) %>%
  var_labels(
    bwt = 'Birth weight (g)',
    smoke = 'Smoking status',
    race = 'Race',
  )
```

```{r}
birthwt %>%
  bar_error(bwt ~ smoke) %>%
  axis_labs()
```

Quick normality check:

```{r}
birthwt %>%
  qq_plot(~ bwt|smoke) %>%
  axis_labs()
```
The (unadjusted) t-test:
```{r}
t.test(bwt ~ smoke, data = birthwt)
```
Final plot with annotation to highlight statistical difference (unadjusted):
```{r}
birthwt %>%
  bar_error(bwt ~ smoke) %>%
  axis_labs() %>%
  gf_star(x1 = 1, x2 = 2, y1 = 3400, y2 = 3500, y3 = 3550, "**")
```
strip_error and bar_error can generate plots stratified by a third variable, for example:
```{r}
birthwt %>%
  bar_error(bwt ~ smoke, fill = ~ Race) %>%
  gf_refine(ggsci::scale_fill_jama()) %>%
  axis_labs()
```

```{r}
birthwt %>%
  bar_error(bwt ~ smoke|Race) %>%
  axis_labs()
```
 analysis of the linear model of smoking on birth weight, adjusting by race (and not by other potential confounders)
```{r}
birthwt %>%
  strip_error(bwt ~ smoke, pch = ~ Race, col = ~ Race) %>%
  gf_refine(ggsci::scale_color_jama()) %>%
  axis_labs()
```
 analysis of the linear model of smoking on birth weight, adjusting by race (and not by other potential confounders)
```{r}
model_bwt <- lm(bwt ~ smoke + race, data = birthwt)

model_bwt %>%
  glm_coef(labels = model_labels(model_bwt))
```

Similar results can be obtained with the function tab_model from the sjPlot package.
```{r}
sjPlot::tab_model(model_bwt,  collapse.ci = TRUE)
```

```{r}
pubh::multiple(model_bwt, ~ race)$df
```


Some advantages of glm_coef over tab_model are:

+ Script documents can be knitted to Word and LaTEX
 (+ (besides HTML).
+ Uses robust standard errors by default. The option to not use robust standard errors is part of the arguments.
+ Recognises some type of models that tab_model does not recognise.

Some advantages of tab_model over glm_coef are:

+ Recognises labels from variables and use those labels to display the table.
+ Includes some statistics about the model.
+ It can display more than one model on the same output.
+ Tables are more aesthetically appealing.

In the previous table of coefficients confidence intervals and p-values for race had not been adjusted for multiple comparisons. We use functions from the emmeans package to make the corrections.

```{r}
multiple(model_bwt, ~ race)$fig_ci %>%
  gf_labs(x = "Difference in birth weights (g)")
```

```{r}
multiple(model_bwt, ~ race)$fig_pval %>%
  gf_labs(y = " ")
```

```{r}
data(Bernard)
head(Bernard)
```

```{r}
Bernard %>%
  mutate(
    fate = relevel(fate, ref = "Dead"),
    treat = relevel(treat, ref = "Ibuprofen")
  ) %>%
  copy_labels(Bernard) %>%
  pubh::cross_tab(fate ~ treat) 
```

```{r}
dat <- matrix(c(84, 140 , 92, 139), 
              nrow = 2, byrow = TRUE)
epiR::epi.2by2(as.table(dat))
```

```{r}
Bernard %>%
  pubh::contingency(fate ~ treat)
```

```{r}
data(oswego, package = "epitools")

oswego <- oswego %>%
  mutate(
    ill = factor(ill, labels = c("No", "Yes")),
    sex = factor(sex, labels = c("Female", "Male")),
    chocolate.ice.cream = factor(chocolate.ice.cream, labels = c("No", "Yes"))
  ) %>%
  var_labels(
    ill = "Developed illness",
    sex = "Sex",
    chocolate.ice.cream = "Consumed chocolate ice cream"
  )
```

```{r}
oswego %>%
  mutate(
    ill = relevel(ill, ref = "Yes"),
    chocolate.ice.cream = relevel(chocolate.ice.cream, ref = "Yes")
  ) %>%
  copy_labels(oswego) %>%
  pubh::cross_tab(ill ~ sex + chocolate.ice.cream) 
```

```{r}
oswego %>%
  pubh::mhor(ill ~ sex/chocolate.ice.cream)
```

```{r}
data(Oncho)


pubh::odds_trend(mf ~ agegrp, data = Oncho, angle = 0, hjust = 0.5)$fig
```
# Diagnostic test

```{r}
Freq <- c(1739, 8, 51, 22)
BCG <- gl(2, 1, 4, labels=c("Negative", "Positive"))
Xray <- gl(2, 2, labels=c("Negative", "Positive"))
tb <- data.frame(Freq, BCG, Xray)
tb <- expand_df(tb)

pubh::diag_test(BCG ~ Xray, data = tb)
```

```{r}
pubh::diag_test2(22, 51, 8, 1739)
```
