---
title: Visualizing the dental workforce of OECD countries I
author: Sergio Uribe
date: '2018-03-08'
slug: visualizing-the-dentists-workforce-of-oecd-countries
categories:
  - visualization
tags:
  - ggplot2
---

The Organisation for Economic Co-operation and Development host a [database](https://data.oecd.org/) with extensive data. In this post we will some visualizations to compare the number of dentists in each country. 

Packages used: 

  - tidyverse
  - gghighlight
  - kableExtra

```{r tidyverse, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
```


First we load the data. Now there is a package ([OECD](https://cran.r-project.org/web/packages/OECD/index.html)) able to extract the datasets, but I will use a local copy: 

```{r dataset}
dent_oecd <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vStv7Pr69DtRKv6Nw6gVBep8hbT3pEeO6B1vNwxK_1DUHgpoTgbuRpZ4SvgtHFQnBZJVGeeQVyRuXZl/pub?gid=1330297229&single=true&output=csv")
```
Always is preferable to take a look the data and its structure: 

```{r head}
head(dent_oecd)
```
```{r glimpse}
glimpse(dent_oecd)
```

And a summary: 

```{r summary}
summary(dent_oecd)
```



I will use the `Dentists graduates`, and `Per 100 000 population` so let's filter and replace the former data set

```{r filter only graduate and 100 000}
dent_oecd <- dent_oecd %>% 
  filter(Variable == "Dentists graduates", 
         Measure == "Per 100 000 population")
```


There are some useless columns, so unselect them

```{r unselect vars}
dent_oecd <- select(dent_oecd, -c(VAR, UNIT, COU, YEA,`Flag Codes`, Flags))
```

Let's plot. Since is a temporal trend, a line plot could be a good idea: 
```{r messy plot}
dent_oecd %>%  # this means: take the dent_oecd dataframe, and
  ggplot(aes(x = Year, y = Value, color = Country)) + #plot the year in the X axis, the value in the y axis and color the lines per country
  geom_line() + 
  labs(y = "Dentists graduation per 100 000 hab")
```

Hmmm....the problem here is that isn't easy to differentiate the countries. 

We have to choices: 
  1. plot each country separate, or
  2. highlight one or few countries from the group we can use the package
  
## 1. Plotting one graph per country: faceting.
```{r facetted plot}
dent_oecd  %>% 
    ggplot(aes(x = Year, y = Value)) + # we delete the color = Country, since is not necessary
  geom_line() + 
  facet_wrap(~Country) + #this means: plot separately each country 
  theme(axis.text.x = element_text(colour="grey20",size=6)) + 
  labs(y = "Dentists graduation per 100 000 hab")
  
```

That's much better. Some observations 

  - seem in general dentist graduating per 100 000 habitants has been stable in OECD countries, 
  - Chile seems to be increasing at a higher rate his graduation rate. Also Lithuania seems to share this trend.
   - The line in Portugal is higher than in the rest of the countries
   - Few countries decrease their graduation rate, seems that Denmark, Finland and Norway share this trend


_TODO: learn how to order the plots, e.g. from higher rate to lower _

We can transform the Value var using the log10. With `ggplot2` we can directly add the transformation to any axis, as y-axis in this case. We use `scale_y_log10()`. Sometimes is worth to use to expand some small differences or to graph data with orders of magnitude of difference. 

```{r facetted plot 2 log10, warning=FALSE}
dent_oecd  %>% 
  filter(Year < 2016) %>% 
    ggplot(aes(x = Year, y = Value)) + 
  geom_line() + 
  facet_wrap(~Country) + 
  theme(axis.text.x = element_text(size=6)) + 
  labs(y = "log10 Dentists graduation per 100 000 hab") + 
  scale_y_log10()
  
```

## 2. Using `gghighlight`
`gghighlight` is a package that highlight ggplot's Lines and Points with Predicates

```{r package ggh, echo=FALSE, results='hide'}
library(gghighlight)
```

Remember the first graph: 

```{r messy plot again}
dent_oecd %>%  
  ggplot(aes(x = Year, y = Value, color = Country)) + 
  geom_line() + 
  labs(y = "Dentists graduation per 100 000 hab")
```

With gghighlight we can, as the name implies, highlight some lines according to a criteria or predicate. 

```{r highlighted and nice line plot}
dent_oecd %>% 
  filter(Year < 2016) %>% 
  gghighlight_line(aes(x = Year, y = Value, color = Country), 
                   predicate = mean(Value) > 5) + # this is the threshold
  labs(y = "Dentists graduation per 100 000 hab") 
```

## Pivot table

Pivot table in the `dplyr` way
```{r}
dent_oecd %>% 
  group_by(Country, Year) %>% 
  summarise(average = mean(Value)) %>% 
  spread(Year, average)
```

Now enhance the table with the package `kableExtra`: 

`install.packages("kableExtra")`

`library(kableExtra)`

```{r, echo=FALSE, results='hide'}
library(kableExtra)
```
Addf the average as reference value


```{r}
dent_oecd_avg <- dent_oecd %>% 
  group_by(Year) %>% 
  summarise(Value = mean(Value, na.omit = TRUE)) # create a new dataframe with the average per year  

dent_oecd_avg[,"Country"] <- "OECD average"  # add a column and fill with a string
dent_oecd <- bind_rows(dent_oecd_avg, dent_oecd) # bind the dataframes by columns
rm(dent_oecd_avg) # remove the average dataset

```

First I create the table and store as an object called `table_dent`
```{r}
table_dent <- dent_oecd %>% 
  group_by(Country, Year) %>% 
  summarise(average = mean(Value)) %>% 
  spread(Year, average) 
```

then apply the `kable` enhance to the object `table_dent` to show a nice formatted table: 

```{r, results='asis'}
knitr::kable(table_dent, 
             caption = "Dentists graduating per 100 000 habs. OECD countries", 
             digits = 1)
```

More information about ggplot2

 - [ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
 - [ggplot2 Search by task](https://www.computerworld.com/article/2935394/business-intelligence/my-ggplot2-cheat-sheet-search-by-task.html)
 - [ggplot 2 reference manual](http://ggplot2.tidyverse.org/reference/)